```{r include=FALSE}
library( ggplot2 )
library( lubridate )
library( readr )
```

# Examples
We have covered enough of the grammar of graphics that you should begin to see the patterns in how it is used to express graphical ideas for `ggplot2`. Here, we will work through some examples.

## Bird Flu mortality
People mail dead birds to the USDA and USGS, where scientists analyze the birds to find out why they died. Right now there is a bird flu epidemic, and the USDA provides public data about the birds in whom the disease has been detected. You can access the data here: [https://www.aphis.usda.gov/aphis/ourfocus/animalhealth/animal-disease-information/avian/avian-influenza/hpai-2022/2022-hpai-wild-birds](https://www.aphis.usda.gov/aphis/ourfocus/animalhealth/animal-disease-information/avian/avian-influenza/hpai-2022/2022-hpai-wild-birds). We will load the data and do some visualization.

```{r}
# load data directly from the USDA website
flu = read_csv( url("https://www.aphis.usda.gov/animal_health/data-csv/hpai-wild-birds-ver2.csv"))
flu$date = mdy(flu$`Date Detected`)

# plot a histogram of when bird flu was detected
ggplot(flu) +
  aes(x=date) +
  geom_histogram() + 
  ggtitle( "Bird flu detections in wild birds") +
  xlab("Date") +
  ylab("Count")

# plot a histogram of when bird flu was detected
ggplot(flu) +
  aes(x=date, fill=`Sampling Method`) +
  geom_histogram() + 
  ggtitle( "Bird flu detections in wild birds") +
  xlab("Date") +
  ylab("Count")

# bar chart shows how the bird flu reports compare between west coast states
subset(flu, State %in% c("California", "Oregon", "Washington")) |>
  ggplot() +
    aes(x=State, fill=`Sampling Method`) +
    stat_count() +
    geom_bar()  + 
    ggtitle( "Bird flu detections in wild birds (West coast states)") +
    ylab("Count")
```


Let's compare the bird flu season to the human flu season. Download hospitalization data for the 2021-2022 and 2022-2023 flu seasons from the CDC website here: [https://gis.cdc.gov/GRASP/Fluview/FluHospRates.html](https://gis.cdc.gov/GRASP/Fluview/FluHospRates.html) (the download button is at the top right and you need to specify which seasons to download).

```{r}
# processing CDC flu data:
cdc = read_csv("data/FluSurveillance_Custom_Download_Data.csv", skip=2)
cdc$date = as_date("1950-01-01")
year(cdc$date) = cdc$`MMWR-YEAR`
week(cdc$date) = cdc$`MMWR-WEEK`


# get flu hospitalization counts that include all race, sex, and age categories
cdc_overall = subset(cdc,
                     `AGE CATEGORY`=="Overall" &
                       `SEX CATEGORY`=="Overall" &
                       `RACE CATEGORY`=="Overall")

# convert the counts to cases per 10 million
cdc_overall$`WEEKLY RATE` = as.numeric(cdc_overall$`WEEKLY RATE`) * 100

# remake the plot but add a new geom_line() with its own data
ggplot(flu) +
  aes(x=date, fill=`Sampling Method`) +
  geom_histogram() +
  geom_line(data=cdc_overall, mapping=aes(x=date, y=`WEEKLY RATE`), inherit.aes=FALSE) +
  ggtitle("Bird flu detections and human flu hospitalizations") +
  xlab("Date") + 
  ylab("Count")
```

## Small business loans
The US Small Business Administration (SBA) maintains data on the loans it offers to businesses. Data about loans made since 2020 can be found at [https://data.sba.gov/dataset/7-a-504-foia](https://data.sba.gov/dataset/7-a-504-foia). We'll load that data and then explore some ways to visualize it.

```{r}
# load the small business loan data
sba = read_csv( url("https://data.sba.gov/dataset/0ff8e8e9-b967-4f4e-987c-6ac78c575087/resource/c71ba6cf-b4e0-4e60-98f0-48aeaf4c6460/download/foia-7afy2020-present-asof-221231.csv"))

# check the SBA data to see the data types, etc.
head(sba)

# boxplot of loan sizes by business type
subset( sba, ProjectState=="CA" ) |>
  ggplot() +
    aes(x=BusinessType, y=SBAGuaranteedApproval) +
    geom_boxplot() +
    scale_y_log10() +
    ggtitle("Small Business Administraton guaranteed loans in California") +
    ylab("Loan guarantee (dollars)")


# relationship between loan size and interest rate
subset( sba, ProjectState=="CA" ) |>
  ggplot() +
    aes(x=GrossApproval, y=InitialInterestRate, ) +
    geom_point() +
    facet_wrap(~BusinessType, ncol=3) +
    scale_x_log10() +
    ggtitle("Interest rate as a function of loan size") +
    xlab("Loan size (dollars)") +
    ylab("Interest rate (%)")
```


Now let's see how we can add aesthetic mappings to a plot in order to convey information on a new dimension. In this case, we will color the points by the loan status. If the default colors aren't working for you, `ggplot2` can integrate directly with [Color Brewer (colorbrewer2.org)](https://colorbrewer2.org). A smoother can help indicate differences that would otherwise be lost in the noise.
```{r}
# color the dots by the oan status.
subset( sba, ProjectState=="CA" & LoanStatus != "EXEMPT" & LoanStatus != "CHGOFF" ) |>
  ggplot() +
    aes(x=GrossApproval, y=InitialInterestRate, color=LoanStatus) +
    geom_point() +
    geom_smooth() +
    facet_wrap(~BusinessType, ncol=3) +
    scale_x_log10() +
    ggtitle("Interest rate as a function of loan size by loan status") +
    xlab("Loan size (dollars)") +
    ylab("Interest rate (%)") + 
    labs( color="Loan status") + 
    scale_color_brewer(type="qual", palette="Accent")
```